---
title: "ESAS preliminary analysis"
author: "Shelly Li"
format: html
execute:
  echo: false
  warning: false
  error: false
tbl-cap-location: top
---

```{r}
#| echo: false
# Load useful libraries
library(tidyverse)
library(readxl)
library(gt)
library(gtsummary)
library(ggplot2)
library(lubridate)
library(psych)
library(scales)
library(texreg)
library(psych)
library(lme4)
```

```{r}
#| echo: false
# Load data - most recent version saved 3/13/2024
# ESAS_raw<-read_excel("J:/Jim_Gonzalez_Research/DATA/Jim/ESAS_datasets/MortonPlant/2023_03_11_ESAS Data.xlsx", na = "N/A")
ESAS_raw<-read_excel("/Volumes/jim_gonzalez_research/DATA/Jim/ESAS_datasets/MortonPlant/2023_03_11_ESAS Data.xlsx", na = "N/A")
# ESAS Data management
ESAS <- ESAS_raw %>%
  mutate(patient_id = as.factor(`Patient ID`)) %>%
  arrange(`Patient ID`, `Date of Visit`) %>%
  group_by(patient_id) %>%
  mutate(date_last_HE = case_when(`Number of Hospitalizations or ED Visits` > 0 ~ 
                                    max(`Date of ED/Hospitalization Visit...45`, 
                                        `Date of ED/Hospitalization Visit...47`,
                                        `Date of ED/Hospitalization Visit...49`,
                                        `Date of ED/Hospitalization Visit...51`,
                                        `Date of ED/Hospitalization Visit...53`,
                                        `Date of ED/Hospitalization Visit...55`,
                                        `Date of ED/Hospitalization Visit...57`,
                                        `Date of ED/Hospitalization Visit...59`,
                                        `Date of ED/Hospitalization Visit...61`,
                                        `Date of ED/Hospitalization Visit...63`,
                                        `Date of ED/Hospitalization Visit...65`,
                                        `Date of ED/Hospitalization Visit...67`,
                                        `Date of ED/Hospitalization Visit...69`, 
                                        na.rm = T), 
                                  TRUE ~ NA), 
         week = interval(first(`Date of Visit`), `Date of Visit`) %/% weeks(1),
         visit = ifelse(`Number of Hospitalizations or ED Visits`>= 1, 1, 0), 
         n_weeks = n(), 
         max_week = max(week)) %>% ungroup()

ESAS_invalid <- ESAS %>%
  filter(is.na(Pain) | `Date of Visit` > date_last_HE) %>%
  select(`Patient ID`, `Date of Visit`, date_last_HE, `Initial Consultation Date`, 
         week, n_weeks, max_week,
         Pain, `Total Score`, `Number of Hospitalizations or ED Visits`)

ESAS_valid <- ESAS %>%
  filter(!is.na(Pain) &
         (is.na(date_last_HE) | `Date of Visit` <= date_last_HE)) %>%
  group_by(patient_id) %>%
  mutate(week = interval(first(`Date of Visit`), `Date of Visit`) %/% weeks(1),
         reading_num = row_number(), 
         visit = ifelse(`Number of Hospitalizations or ED Visits`>= 1, 1, 0), 
         n_weeks = n(), 
         max_week = max(week)) %>%
  ungroup() 
write_csv(ESAS_valid, "/Volumes/jim_gonzalez_research/DATA/Jim/ESAS_datasets/MortonPlant/Analysis_MP.csv")
```

## Preliminary findings from ESAS data - Morton Plant 
* In raw data set, there are `r sum(!is.na(ESAS$patient_id))` observations, and `r n_distinct(ESAS$patient_id)` patients.

* We cleaned the data set, and included visits happened before the last hospitalization or ED visit and when there are valid ESAS scores. After data cleanning, we have `r sum(!is.na(ESAS_valid$patient_id))` observations, and `r n_distinct(ESAS_valid$patient_id)` patients.

### Spaghetti Plot of ESAS Data
```{r}
#| echo: false
#| warning: false
ESAS_valid %>%
  ggplot(aes(x = week, y = `Total Score`, group = patient_id, color = patient_id)) +
  geom_line() +
#  geom_point() +
  labs(title = "", x = "Week", y = "ESAS Total Score") +
  theme_classic() +
  theme(legend.position = "bottom") +
    guides(color = guide_legend(nrow = 5, title = NULL, title.position = "top"))
```

```{r}
#| echo: false
# Using by() function
patient_coefficients <- by(ESAS_valid, ESAS_valid$patient_id, function(subset) {
  lm_model <- lm(`Total Score` ~ week, data = subset)
  intercept <- coef(lm_model)[1]
  slope <- coef(lm_model)[2]
  return(data.frame(patient_id = unique(subset$patient_id),
                    intercept = intercept,
                    slope = slope))
})

# Combine the results into a single data frame
patient_coefficients_merge <- do.call(rbind, patient_coefficients) %>%
  merge(ESAS_valid, by = "patient_id") %>%
  filter(week == 0) %>%
  select(patient_id, intercept, slope, visit, n_weeks, max_week,
         `Number of Hospitalizations or ED Visits`, 
         Sex, Race, Ethnicity, `BMI (kg/m^2)`, `Marital Status`) %>%
  mutate(slope_fill = ifelse(is.na(slope), 0, slope), 
         intercept_fill = ifelse(is.na(slope), mean(intercept), intercept))
# View the results
# print(patient_coefficients)
```

### Span of data collection in weeks for all observations (n = `r sum(!is.na(ESAS$patient_id))`):
```{r}
#|label: tbl-weeks
describe(ESAS_valid$week) %>%
  as.data.frame() %>%
  select(-vars, - trimmed, -mad) %>%
  gt() %>%
  cols_align(align = "center")%>%
  fmt_number(c(mean, sd, median, min, max, range, skew, kurtosis, se), decimals = 1)%>%
  tab_options(table.width = "100%")
```

### Span of data collection in weeks for all participants (n = 53)
```{r}
#|label: tbl-maxweek
describe(patient_coefficients_merge$max_week) %>%
  as.data.frame() %>%
  select(-vars, - trimmed, -mad) %>%
  gt() %>%
  cols_align(align = "center")%>%
  fmt_number(c(mean, sd, median, min, max, range, skew, kurtosis, se), decimals = 1)%>%
  tab_options(table.width = "100%")
```

```{r}
#| echo: false
patient_coefficients_merge %>%
  ggplot(aes(x = max_week)) +
  geom_histogram(binwidth = 1, color = "black", fill = "skyblue", alpha = 0.7) +
  labs(title = "Histogram of last follow-up (n = 53)", x = "Last follow-up week", y = "Frequency") +
  theme_minimal()
```

### Number of time points for all participants (n = 53)
```{r}
#|label: tbl-numweek
describe(patient_coefficients_merge$n_weeks) %>%
  as.data.frame() %>%
  select(-vars, - trimmed, -mad) %>%
  gt() %>%
  cols_align(align = "center")%>%
  fmt_number(c(mean, sd, median, min, max, range, skew, kurtosis, se), decimals = 1)%>%
  tab_options(table.width = "100%")
```

```{r}
#| echo: false
patient_coefficients_merge %>%
  ggplot(aes(x = n_weeks)) +
  geom_histogram(binwidth = 1, color = "black", fill = "skyblue", alpha = 0.7) +
  labs(title = "Histogram of number of time points (n = 53)", x = "Number of time points", y = "Frequency") +
  theme_minimal()
```

```{r}
#| echo: false
#| results: asis
#| tbl-cap-location: top
patient_coefficients_merge %>%
  mutate(value_time = ifelse(n_weeks == 1, "Baseline only", "More than one time")) %>%
  select(-patient_id, -slope_fill, -intercept_fill) %>%
#  mutate(n_weeks = as.factor(n_weeks))%>%
  tbl_summary(by = value_time) %>%
  add_overall() %>%
  modify_caption("**Table 1:** Sample characteristics") 
```
Note: There are 2 cases that are missing hospitalization info. 

### Logistic regressions of baseline and change (slope) ESAS value on hospitalizations or ED visits (Y/N). 
```{r}
#| results: asis
model_raw_int <- glm(visit ~ intercept, data = patient_coefficients_merge, family = binomial)
model_raw_sl <- glm(visit ~ slope, data = patient_coefficients_merge, family = binomial)
model_int_sl <- glm(visit ~ intercept + slope, data = patient_coefficients_merge, family = binomial)
```

```{r}
#|label: tbl-weeks
#|tbl-cap: Logistic regressions of baseline ESAS value on hospitalizations or ED visits (Y/N) (n = 51).
#| results: asis
model_raw_int %>%
  tbl_regression(exponentiate = T,
                 label = list(intercept = "Baseline ESAS")) %>%
  modify_caption("**Table 2:** Logistic regressions: baseline ESAS value on hospitalizations or ED visits (Y/N) (n = 51)")
```

```{r}
#| results: asis
model_raw_sl %>%
  tbl_regression(exponentiate = T, 
                 label = list(slope = "Change of ESAS"))%>%
  modify_caption("**Table 3:** Logistic regressions: slope of ESAS value on visits (Y/N) (n = 29)")
```

```{r}
#| results: asis
model_int_sl %>%
  tbl_regression(exponentiate = T, 
                 label = list(intercept = "Baseline ESAS", 
                              slope = "Change of ESAS"))%>%
  modify_caption("**Table 4:** Logistic regressions: baseline and slope ESAS value on visits (Y/N) (n = 29)")
```