---
title: "ESAS preliminary analysis"
author: "Shelly Li"
format: pdf
execute:
  echo: false
  warning: false
  error: false
---

## Preliminary findings from ESAS data - Morton Plant 
### Current problems: 
#### 1. ! Do we remove ESAS values that is too far away from baseline?
#### 2. SOme patients have more than one ED/hospitalization visit dates - do we only analyze the number of visits? 


```{r}
#| echo: false
# Load useful libraries
library(tidyverse)
library(readxl)
library(gt)
library(gtsummary)
library(ggplot2)
library(lubridate)
library(psych)
library(scales)
library(texreg)
library(psych)
library(lme4)
```

```{r}
#| echo: false
# Load data - most recent version saved 3/13/2024
# ESAS_raw<-read_excel("J:/Jim_Gonzalez_Research/DATA/Jim/ESAS_datasets/MortonPlant/2023_03_11_ESAS Data.xlsx", na = "N/A")
ESAS_raw<-read_excel("/Volumes/jim_gonzalez_research/DATA/Jim/ESAS_datasets/MortonPlant/2023_03_11_ESAS Data.xlsx", na = "N/A")
# ESAS Data management
ESAS <- ESAS_raw %>%
  mutate(patient_id = as.factor(`Patient ID`)) %>%
  arrange(`Patient ID`, `Date of Visit`) %>%
  group_by(`Patient ID`) %>%
  mutate(week = interval( first(`Date of Visit`), `Date of Visit`) %/% weeks(1),
         reading_num = row_number())
```

There are `r sum(!is.na(ESAS$patient_id))` observations, and `r n_distinct(ESAS$patient_id)` patients. 

The following are the destribution of the weeks 
```{r}
#|label: tbl-weeks
describe(ESAS$week) %>%
  as.data.frame() %>%
  select(-vars, - trimmed, -mad) %>%
  gt() %>%
  cols_align(align = "center")%>%
  fmt_number(c(mean, sd, median, min, max, range, skew, kurtosis, se), decimals = 1)%>%
  tab_options(table.width = "100%")
```

### Histogram with blocks of 4 weeks - Remove 
```{r}
#| echo: false
ESAS %>%
  ggplot(aes(x = week)) +
  geom_histogram(binwidth = 4, color = "black", fill = "skyblue", alpha = 0.7) +
  labs(title = "Histogram with Blocks of 4 Weeks", x = "Week", y = "Frequency") +
  theme_minimal()
```

### Plot spaghetti plot
```{r}
#| echo: false
#| warning: false
ESAS %>%
  ggplot(aes(x = week, y = `Total Score`, group = patient_id, color = patient_id)) +
  geom_line() +
#  geom_point() +
  labs(title = "Spaghetti Plot of ESAS Data", x = "Week", y = "ESAS Total Score") +
  theme_classic() +
  theme(legend.position = "bottom") +
    guides(color = guide_legend(nrow = 5, title = NULL, title.position = "top"))
  
```